Distinguishing multiple-merger from Kingman coalescence using two-site frequency spectra

### 2018-08-29 manuscript/manuscript.tex
Make genetic distance notation consistent.

For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with total recombination rate, $r$ (See SI for parameter combinations).
We chose values of $r$ so that the realized values of $r\E{T_2}$ varied over several orders of magnitude, where $\E{T_2}$ was measured from the simulation output.
For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
These expectations allow us to calculate all of the 2-SFS statistics defined above.

@@

\Fig{fig:wfpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
In both models, $\fpmi > 0$ for $i\approx j$ for $r\E{T_2} \sim 1$.
This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our binned $\hilopmi$.
On the other hand, the positive $\fpmi$ in the beta coalescent persists for $r\E{T_2} > 1$.
Thus, the signal of multiple mergers in $\fpmi$ is robust to recombination.

@@

\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d=r\E{T_2}$ and three different choices of $i_c$.
Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS, which is independent of the recombination rate.
At low recombination rates, $\hilopmi$ may be greater in growing populations than in constant-$N$ populations (dotted lines), but is always less than zero, consistent with the results for non-recombining sites.
When $r\E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
In all Kingman models, $\hilopmi$ decays to zero for $r\E{T_2}\gg 1$.

@@

\Fig{fig:hilo_vs_tajD} shows the relative invariance of $\hilopmi$ to a time-varying coalescent rate.
While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $r \E{T_2} \sim 1$ apart.
This holds for exponential as well as piecewise constant growth.
In contrast, the beta coalescent induces large positive $\hilopmi$ for the same values of Tajima's D.
The figure also showes that this result is robust to the choice of cutoff minor allele count for binning.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD_recombination.pdf}
\caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus normalized Tajima's D for simulations with recombination. All values are for $r \E{T_2} \sim 1$ and $n=100$. Orange points show Kingman coalescent simulations with population growth. Blue points show beta coalescent simulations. [Q: Should this be the bottom row of the previous figure?] \label{fig:hilo_vs_tajD}}
\end{figure}

@@

The first row of \fig{fig:dpgp3} shows that the expected SFS under the fit demographic models agree with the observed SFS, demonstrating that a time-varying $N(t)$ can explain this aspect of the data well.
In contrast, the second row shows the expected and observed weighted $\fpmi$ for averaged over distances less than $15 d_c$ apart, where $d_c = 2/ (\Pi \mu)$ is the distance scale corresponding to $r\E{T_2} = 1$.
Here, the data shows strong positive associations between nearby alleles at different frequencies, while the model of population growth predicts weak negative associations except just off of the diagonal.
This pattern extends across a range of genomic distances, $d/d_c \in (10^{-1}, 10^2)$ (\fig{fig:dpgp3}, third row).
As a result, we may conclude that the data is not well explained by the Kingman coalescent with population growth.

>>>>

For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per basepair recombination rate $r$ resulting in a per locus total recombination rate $d r$.
(See SI for parameter combinations.)
We chose values of $d r$ so that the realized values of $d r \E{T_2}$ varied over several orders of magnitude.
For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
These expectations allow us to calculate all of the 2-SFS statistics defined above.

@@

\Fig{fig:wfpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
In both models, $\fpmi > 0$ for $i\approx j$ for $d r \E{T_2} \sim 1$.
This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our binned $\hilopmi$.
On the other hand, the positive $\fpmi$ in the beta coalescent persists for $d r \E{T_2} > 1$.
Thus, the signal of multiple mergers in $\fpmi$ is robust to recombination.

@@

\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d r \E{T_2}$ and three different choices of $i_c$.
Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS, which is independent of the recombination rate.
At low recombination rates, $\hilopmi$ may be greater in growing populations than in constant-$N$ populations (dotted lines), but is always less than zero, consistent with the results for non-recombining sites.
When $d r \E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
In all Kingman models, $\hilopmi$ decays to zero for $d r \E{T_2}\gg 1$.

@@

\Fig{fig:hilo_vs_tajD} shows the relative invariance of $\hilopmi$ to a time-varying coalescent rate.
While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $d r \E{T_2} \sim 1$ apart.
This holds for exponential as well as piecewise constant growth.
In contrast, the beta coalescent induces large positive $\hilopmi$ for the same values of Tajima's D.
The figure also showes that this result is robust to the choice of cutoff minor allele count for binning.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD_recombination.pdf}
\caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus normalized Tajima's D for simulations with recombination. All values are for $d r \E{T_2} \sim 1$ and $n=100$. Orange points show Kingman coalescent simulations with population growth. Blue points show beta coalescent simulations. [Q: Should this be the bottom row of the previous figure?] \label{fig:hilo_vs_tajD}}
\end{figure}

@@

The first row of \fig{fig:dpgp3} shows that the expected SFS under the fit demographic models agree with the observed SFS, demonstrating that a time-varying $N(t)$ can explain this aspect of the data well.
In contrast, the second row shows the expected and observed weighted $\fpmi$ for averaged over distances less than $15 d_c$ apart, where $d_c = 2/ (\Pi \mu)$ is the distance scale corresponding to $d_c r \E{T_2} = 1$.
Here, the data shows strong positive associations between nearby alleles at different frequencies, while the model of population growth predicts weak negative associations except just off of the diagonal.
This pattern extends across a range of genomic distances, $d/d_c \in (10^{-1}, 10^2)$ (\fig{fig:dpgp3}, third row).
As a result, we may conclude that the data is not well explained by the Kingman coalescent with population growth.
